Subset coercions in \Coq

### 2007-01-31 papers/article/main.tex
Before strip for sending

\documentclass{llncs}
\usepackage{abbrevs}
%\usepackage{array}
\usepackage{alltt}
\def\text{\mbox}

%%%% FixMe: remove 
\pagestyle{plain}

@@

\institute{LRI, Paris Sud XI University\\\email{sozeau@lri.fr}}
\title{Subset coercions in \Coq} % \\ \small{Draft - \today}}
\date{\today}

@@

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpit$}}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpip$}}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{-1em}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset} 
  \vspace{-1em}
\end{figure}

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{-1em}
  \caption{\CIC typing judgement}
  \label{fig:cic-decl-rules}
  \vspace{-1em}
\end{figure}

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{-1em}
  \caption{\Russell new rules}
  \label{fig:typing-decl-rules}
  \vspace{-1em}
\end{figure}

@@

Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A. B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : unit `| A \}. B$ (with \verb|unit| the
type containing a single element \verb|tt|) to benefit from
the subtyping mechanism which we will define later.

The reduction rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\downarrow{x}$ the complete normal form
of $x$ and by $\hnf{x}$ its weak head normal form. We use a standard
judgemental equality $`G \type T \eqbp U : s$.

@@

\begin{figure}[h]
  \vspace{-1em}
  \subtdRules
  \vspace{-1em}
  \caption{\Russell{} conversion}
  \label{fig:conversion-decl-rules}
  \vspace{-1em}
\end{figure}

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgement. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubTrans} rule ensure
that our judgement builds an \emph{equivalence} and has
proper modularity. It is trivial to check that we have symmetry given 
the symmetry of the definitional equality. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

%     \vspace{\infvspace}
%     \SubsetR\DP
  \end{center}
  \vspace{-1em}
  \caption{\Russell{} algorithmic typing, new rules}
  \label{fig:typing-algo-rules}
%  \vspace{-1em}
\end{figure*}

@@

    \vspace{\infvspace}
    \SubSubAs\DP
  \end{center}
  \vspace{-1em}
  \caption{\Russell{} algorithmic equivalence}
  \label{fig:subtyping-algo-rules}
  \vspace{-1em}
\end{figure*}

@@

  We denote by $\typeafn{`G}{t}$ the function which returns the type of
  term $t$ in context $`G$ such that $`G \typea t : \typeafn{`G}{t}$.  
\end{paragraph}

@@

\subsection{From \CIC to \lng{}}
We have presented a calculus based on \CIC with a stronger equivalence
but with restricted typing rules. We will now see how they relate
formally.
We can build a forgetful map from terms of \CIC to \Russell
(interpreting inductives as constants). It bears
relation to the $`e$ extraction function defined by Werner in
\cite{Werner:ProofIrrelevance}. Essentially, \Russell{} terms do not
contain or manipulate logical information attached to objects of subset
types, hence we must forget about it when translating.

We define the forgetful map $()^{\circ}$ as an homomorphism on terms
except for the following cases:
\begin{center}
  \CCtoRussellforgetFig
\end{center}

@@

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms as they do not contain the constructor
$\id{elt}$ or subset projections. Moreover, this erasure function will be
defined on all \Coq terms elaborated from \Russell.
Indeed, the second projection of a
subset element will only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it does not exist.
Indeed, in \PVS{} you cannot manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$. If we allowed the second projection in \Russell, that would be doable
and we would have an inconsistency.
Hence we left only the subset type forming rule (figure
\vref{fig:typing-decl-rules}) in \Russell, while the introduction and
first projection of subset elements are internalized by the coercion judgement.

@@

\begin{center}
  \vspace{-1em}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
  \DP
  \vspace{-1em}
\end{center}
We restrict it to objects of type $\Prop$ because we consider that the
informative part of the algorithm has been entirely given in the
original term.

@@

The proof of this proposition is the main technical contribution of this
paper. It also represented the main difficulty in our work. 
The remainder of this section is organised as follows: first we 
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we define the
interpretation of terms which is mutually recursive with the previous
algorithm. Finally we present a proof sketch of the aforementioned
theorem.

@@

\begin{figure}[ht]
  \vspace{-1em}
  \subtisc
  \vspace{-1em}
  \caption{Coercion derivation}
  \label{fig:coerce-impl-rules-short}
  \vspace{-1em}
\end{figure}

Coercions are formalised as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(i.e., it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and adapted for checking coherence 
of coercions, which we do not need here as we can get uniqueness
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables and depend on a
typing context. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respects some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this was not adapted to our setting, so we invented this
new presentation.

@@

\subsection{Interpretation of terms}
The interpretation of terms from \Russell to \CICq (figure
\vref{fig:interp}) is a straightforward
recursive traversal of algorithmic typing derivations. In other, more
practical words, we can do interpretation of terms simultaneously with
typing in our implementation, just like for the existing coercion system
of \Coq.

\begin{figure}[ht]
  \vspace{-1em}
  \interparr
  \vspace{-1em}
  \caption{Interpretation of terms}
  \label{fig:interp}
  \vspace{-1em}
\end{figure}

We describe the case of application here; others are similar. First we
get the \Russell types $F$ and $U$ of the function $f$ and argument $u$ by calling the
typing function on both terms. Then we must ensure that $F$ can indeed
be seen as a product using the $\mualgo$ operator. We can build
the coercion between $F$ and this product and between $U$ and its
domain. We finally instantiate these coercions by their corresponding
interpreted objects and return their application.

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms into a system with only
$\beta\pi$-equivalence and still be a simple extension of \CIC. 
Indeed, consider the following example. Suppose we have two types $A, B$
with coercions: $c : A \suba B$ and $d : B \suba A$. Additionally, let
$P : A "->" Prop$ with introduction terms $p : \Pi x : A, P~x$ and $q :
\Pi x : B, P~x$. Consider the following well-typed \Russell term in
context $`G = x : A$: $p~x =_{P~x} q~x$. By interpretation into \CIC, we
have $\ip{P}{`G} = P$, $\ip{p}{`G} = p$ and $\ip{q}{`G} = \lambda x :
B. q~d[x] : \Pi x : B, P~d[x]$. Now the interpretation of the equality
gives:
$p~x =_{P~x} \ip{q~x}{`G}$ where $\ip{q~x}{`G} `= (\lambda x : B. q~d[x]) c[x] `=
q~d[c[x]]$. The right hand side has type $P~d[c[x]]$ which must be
convertible with $P~x$ for the whole term to be well-typed. This
indicates that coercions must be unique (so that $d `o c `= id_A `=
\ctxdot$), which implies that we have to include $\eta$ 
conversion in the definitional equality. For
example if $A = \Pi x : X.Y$ and $B = \Pi x : X'.Y'$ with $X = X'$ and
$Y = Y'$, we have $d `o c = \lambda x : X. \ctxdot~x `o \lambda x
: X'. \ctxdot~x = \lambda x : X. \ctxdot~x$ which must be definitionally
equal to $\ctxdot$.
Similarly, we must add $\eta$ rules for all introduction terms of our
language to ensure uniqueness of coercions. It is an important
property of our system, as it corresponds in practice to the determinism
of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects ($\tau$) and is sufficient to ensure uniqueness of coercions.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the Calculus of Constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{-1em}
  \ccqeqarr
  \vspace{-1em}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{-1em}
\end{figure}

@@

\subsection{Further work}
Much has to be done to smoothly integrate \Russell into \Coq. The
treatment of existential variables of \Coq has to be improved, the
type inference algorithm needs some tuning to become more similar to \ML
and an integration of proof-irrelevance in the kernel, while rarely needed in
practice, is necessary to have a robust system.
We also hope to extend this mechanism of proof-obligations generation to other
constructs, notably inductive types. Finally, we intend to use this system as a
basis for interpretation of complete \ML{} programs in \Coq, using a monadic
translation or some kind of effects to reflect imperative constructs.

>>>>

\documentclass{llncs}
\usepackage{abbrevs}
\usepackage{alltt}
\def\text{\mbox}

\def\figureshrink{0em}

@@

\institute{Universit\'e Paris Sud, CNRS, Laboratoire LRI, UMR 8623,
  Orsay, F-91405\\
  INRIA Futurs, ProVal, Parc Orsay Universit\'e, F-91893
  \\\email{sozeau@lri.fr}}
\title{Subset coercions in \Coq} % \\ \small{Draft - \today}}
\date{\today}

@@

    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpit$}}
    \UIC{$`G \type \eltpit~t : A$}
    \DP
    \quad
    \AXC{$`G \type t : \mysubset{x}{A}{P}$}
    \LeftLabel{\irule{Subset-$\eltpip$}}
    \UIC{$`G \type \eltpip~t : P[\eltpit~t/x]$}
    \DP
  \end{center}
  \vspace{\figureshrink}
  \caption{Subset type in \CIC}
  \label{fig:cci-subset} 
  \vspace{\figureshrink}
\end{figure}

@@

    \vspace{\infvspace}
    \Conv\DP
  \end{center}
  \vspace{\figureshrink}
  \caption{\CIC typing judgement}
  \label{fig:cic-decl-rules}
  \vspace{\figureshrink}
\end{figure}

@@

    \vspace{\infvspace}
    \SubsetR\DP
  \end{center}
  \vspace{\figureshrink}
  \caption{\Russell new rules}
  \label{fig:typing-decl-rules}
  \vspace{\figureshrink}
\end{figure}

@@

Following the presentation as a Pure Type System, the set of sorts
$\PTSsorts$ is defined as $\setproptype$. As usual, we let $s, s_i$ for
$i `: `N$ range over sorts. The axioms are $\PTSaxioms = \{ (\Set,
\Type), (\Prop, \Type) \}$ and the set of rules $\PTSrules$ is defined
by the functional relation $`A s_1 s_2, (s_1, s_2, s_2) `: \PTSrules$. We allow products $\Pi x :
A. B$ where $A : \Prop$ and $B : \Set$ but the user is encouraged to
encode them as $\Pi x : \{ \_ : \texttt{unit} `| A \}. B$ (with \texttt{unit} the
type containing a single element \verb|tt|) to benefit from
the subtyping mechanism which we will define later.

The reduction rules of this system are the standard $\beta$ and $\pi$
for projections. We denote by $\hnf{x}$ the weak head normal form
of $x$. We use a standard judgemental equality $`G \type T \eqbp U : s$
meaning $T, U$ are well-typed, convertible terms of sort $s$ in context $`G$.

@@

\begin{figure}[h]
  \vspace{\figureshrink}
  \subtdRules
  \vspace{\figureshrink}
  \caption{\Russell{} conversion}
  \label{fig:conversion-decl-rules}
  \vspace{\figureshrink}
\end{figure}

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgement. We use a judgemental equality here, which will be refined by
the usual conversion relation later. The \irule{SubTrans} rule ensures
that our judgement builds an \emph{equivalence} and has
proper modularity. It is trivial to check that we have symmetry given 
the symmetry of the definitional equality. The next two rules (\irule{SubProd} and
\irule{SubSigma}) do context closure for dependent
products and sums. It is remarkable that we use contravariance for
% \cite{journals/toplas/Castagna95}
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symmetry. However, it will
become important when we create coercions (see figure
\vref{fig:coerce-impl-rules-short}).

@@

%     \vspace{\infvspace}
%     \SubsetR\DP
  \end{center}
  \vspace{\figureshrink}
  \caption{\Russell{} algorithmic typing, new rules}
  \label{fig:typing-algo-rules}
%  \vspace{\figureshrink}
\end{figure*}

@@

    \vspace{\infvspace}
    \SubSubAs\DP
  \end{center}
  \vspace{\figureshrink}
  \caption{\Russell{} algorithmic equivalence}
  \label{fig:subtyping-algo-rules}
  \vspace{\figureshrink}
\end{figure*}

@@

  We denote by $\typeafn{`G}{t}$ the function which returns the type of
  term $t$ in context $`G$ such that $`G \typea t : \typeafn{`G}{t}$,
  when it exists.
\end{paragraph}

@@

\subsection{From \CIC to \lng{}}
We have presented a calculus based on \CIC with a stronger equivalence
but with restricted typing rules. We will now see how \Russell and \CIC relate
formally.
We can build a forgetful map from terms of \CIC to \Russell
(interpreting inductives as constants). It bears
relation to the $`e$ extraction function defined by Werner in
\cite{Werner:ProofIrrelevance}. Essentially, \Russell{} terms do not
contain or manipulate logical information attached to objects of subset
types, hence we must forget about it when translating.

We define the forgetful map $()^{\circ}$ from \CIC terms to \Russell
terms plus a distinguished object $`_$ as an homomorphism on terms except for the following cases:
\begin{center}
  \CCtoRussellforgetFig
\end{center}

@@

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms as they do not contain the constructor
$\id{elt}$ or subset projections. Moreover, this erasure function will be
defined on all \Coq terms elaborated from \Russell.
Indeed, the second projection of a
subset element will only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it does not exist. In fact, as we can derive the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$ in \Russell, allowing the second projection would permit to derive
$0 \neq 0$, an inconsistency.
Hence we left only the subset type forming rule (figure
\vref{fig:typing-decl-rules}) in \Russell, while the introduction and
first projection of subset elements are internalized by the coercion judgement.

@@

\begin{center}
  \vspace{\figureshrink}
  \UAX{}
  {$`G \typeq P : \Prop$}
  {$`G \typeq ?_{P} : P$}
  {}
  \DP
  \vspace{-0.2em}
\end{center}
We restrict it to objects of type $\Prop$ because we consider that the
informative part of the algorithm has been entirely given in the
original term.

@@

The proof of this proposition is the main technical contribution of this
work, it also represented the main difficulty. 
The remainder of this section is organised as follows: first we 
define the explicit coercion derivation algorithm which will permit to
put proof obligations in the terms, then we define the
interpretation of terms which is mutually recursive with the previous
algorithm. Finally we present a proof sketch of the aforementioned
theorem.

@@

\begin{figure}[ht]
%  \vspace{\figureshrink}
  \subtisc
%  \vspace{\figureshrink}
  \caption{Coercion derivation}
  \label{fig:coerce-impl-rules-short}
%  \vspace{\figureshrink}
\end{figure}

Coercions are formalised as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instantiation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(i.e., it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied.
% Other
% presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
% was adopted by Gang Chen in his treatment of coercive subtyping for the
% Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
% of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
% in defining a judgement $x : A \type y : B$ which reads: $A$ is
% coercible to $B$ and the coercion is $\lambda x : A.y$. This
% presentation is very elegant and adapted for checking coherence 
% of coercions, which we do not need here as we can get uniqueness
% of coercions. Moreover, it did not seem readily usable in
% our setting, where coercions may contain metavariables and depend on a
% typing context. Similarly, Amokrane Sa\"ibi defines
% a coercion derivation algorithm which respects some coherence criteria,
% roughly, that type-erased coercions $\beta\eta$-reduce to the
% identity. Again, this was not adapted to our setting, so we invented this
% new presentation.

@@

\subsection{Interpretation of terms}
The interpretation of terms from \Russell to \CICq (figure
\vref{fig:interp}) is a straightforward
recursive traversal of algorithmic typing derivations. In other, more
practical words, we can do interpretation of terms simultaneously with
typing in our implementation, just like for the existing coercion system
of \Coq. 
\begin{figure}[ht]
  \vspace{\figureshrink}
  \interparr
  \vspace{\figureshrink}
  \caption{Interpretation of terms}
  \label{fig:interp}
  \vspace{\figureshrink}
\end{figure}
We describe the case of application here; others are similar. First we
get the \Russell types $F$ and $U$ of the function $f$ and argument $u$ by calling the
typing function on both terms. Then we must ensure that $F$ can indeed
be seen as a product using the $\mualgo$ operator. We can build
the coercion between $F$ and this product and between $U$ and its
domain. We finally instantiate these coercions by their corresponding
interpreted objects and return their application.
Well-typedness of a term in \Russell is a sufficient condition for
$\ip{t}{}$ to be defined.

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms into a system with only
$\beta\pi$-equivalence and still be a simple extension of \CIC. 
Indeed, consider the following example. Suppose we have two types $A, B$
with coercions: $c : A \suba B$ and $d : B \suba A$. Additionally, let
$P : A "->" Prop$ with introduction terms $p : \Pi x : A, P~x$ and $q :
\Pi x : B, P~x$. Consider the following well-typed \Russell term $p~x
=_{P~x} q~x$ in context $`G = x : A$. By interpretation into \CIC, we
have $\ip{P}{`G} = P$, $\ip{p}{`G} = p$ and $\ip{q}{`G} = q : \Pi x : B,
P~d[x]$. Now the interpretation of the equality gives:
$p~x =_{P~x} \ip{q~x}{`G}$ where $\ip{q~x}{`G} `= q~c[x]$. 
The right hand side has type $P~d[c[x]]$ which must be
convertible with $P~x$ for the whole term to be well-typed. This
indicates that coercions must be unique (so that $d `o c `= id_A `=
\ctxdot$), which implies that we have to include $\eta$ 
conversion in the definitional equality. For
example if $A = \Pi x : X.Y$ and $B = \Pi x : X'.Y'$ with $X = X'$ and
$Y = Y'$, we have $d `o c = \lambda x : X. \ctxdot~x `o \lambda x
: X'. \ctxdot~x = \lambda x : X. \ctxdot~x$ which must be definitionally
equal to $\ctxdot$.
Similarly, we must add $\eta$ rules for all introduction terms of our
language to ensure uniqueness of coercions. It is an important
property of our system, as it corresponds in practice to the determinism
of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects ($\tau$) and is sufficient to ensure uniqueness of coercions.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the Calculus of Constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{\figureshrink}
  \ccqeqarr
  \vspace{\figureshrink}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{\figureshrink}
\end{figure}

@@

\subsection{Further work}
Much has to be done to smoothly integrate \Russell into \Coq. The
treatment of existential variables of \Coq has to be improved, the
type inference algorithm needs some tuning to become more similar to \ML
and an integration of proof-irrelevance in the kernel, while rarely needed in
practice, is necessary to have a robust system.
We also hope to extend this mechanism of proof-obligations generation to other
constructs, notably (co-)inductive types. Finally, we intend to use this system as a
basis for interpretation of more general \ML{} programs in \Coq, using a monadic
translation or some kind of effects to reflect imperative constructs
like exceptions and references.
